spinifex: An R Package for Creating a Manual Tour of Low-dimensional Projections of Multivariate Data

### 2020-08-25 paper/spyrison-cook.Rmd
Code touch up, gifs saved to /output/.

---
title: > 
  spinifex: An R Package for Creating a Manual Tour of Low-dimensional Projections of Multivariate Data
author:
  - name: Nicholas Spyrison
    affiliation: Monash University 
    address:
    - Faculty of Information Technology
    email:  nicholas.spyrison@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
abstract: >
  Dynamic low-dimensional linear projections of multivariate data collectively known as *tours* provide an important tool for exploring multivariate data and models. The R package \pkg{tourr} provides functions for several types of tours: grand, guided, little, local and frozen. Each of these can be viewed dynamically, or saved into a data object for animation. This paper describes a new package, \pkg{spinifex}, which provides a manual tour of multivariate data where the projection coefficient of a single variable is controlled. The variable is rotated fully into the projection, or completely out of the projection. The resulting sequence of projections can be displayed as an animation, with functions from either the \pkg{plotly} or \pkg{gganimate} packages. By varying the coefficient of a single variable, it is possible to explore the sensitivity of structure in the projection to that variable. This is particularly useful when used with a projection pursuit guided tour to simplify and understand the solution. The use of the manual tour is applied particle physics data to illustrate the sensitivity of structure in a projection to specific variable contributions.
keywords: exploratory data analysis, projection pursuit, high dimensional data, data visualization, statistical graphics, data science
preamble: >
  % preamble packages
link-citations: true
fig_caption: yes
toc: false
bibliography: spyrison-cook.bib
output: rticles::rjournal_article
  #bookdown::pdf_book ## alternative
---
```{r setup_paper, include=FALSE}
knitr::opts_chunk$set(
  fig.align  = "center", 
  echo       = TRUE, 
  collapse   = TRUE,
  message    = FALSE, 
  warning    = FALSE, 
  error      = FALSE,
  cache      = FALSE, # needed for log and figure out?
  cache.lazy = FALSE
)
if (F){ ## From reproducing, please use this dev ver:
  remotes::install_github("nspyrison/spinifex", 
                          ref = "da75976588dc008e21f01ce6a57d9778d0ed6933")
}
library(tourr)
library(spinifex) ## Use dev version listed above!
library(ggplot2)
library(gridExtra)
library(dplyr)
```

@@

- $\textbf{X}$, the data, an $n \times p$ numeric matrix to be projected to 2D.
- $\textbf{B} = (B_1,~ B_2)$, any 2D orthonormal projection basis, $p \times 2$ matrix, describing the projection from $\mathbb{R}^p \Rightarrow \mathbb{R}^2$. This is called this the "original projection" because it is the starting point for the manual tour.
- $k$, is the index of the variable to manipulate, called the "manip var".
<!--- $|\textbf{B}_k|$, the norm of the manip var, $k$, the length of this basis vector between 0 and 1 inclusive.
-->
- $\textbf{e}$, a 1D basis vector of length $p$, with 1 in the $k$-th position and 0 elsewhere.
- $\textbf{M}$ is a $p \times 3$ matrix, defining the 3D subspace where data rotation occurs and is called the manip(ulation) space.
- $\textbf{R}$, the 3D rotation matrix, for performing unconstrained 3D rotations within the manip space, $\textbf{M}$.
- $\theta$, the angle of in-projection rotation, for example, on the reference axes; $c_\theta, s_\theta$ are its cosine and sine.
- $\phi$, the angle of out-of-projection rotation, into the manip space; $c_\phi, s_\phi$ are its cosine and sine. The initial value for animation purposes is $\phi_1$.
- $\textbf{U}$, the axis of rotation for out-of-projection rotation orthogonal to $\textbf{e}$.
- $\textbf{Y}$, the resulting projection of the data through the manip space, $\textbf{M}$, and rotation matrix, $\textbf{R}$.

@@

```{r step0, echo=F, results='hide', out.width='70%',                        fig.cap = "Initial 2D projection: representation of the basis (left) and resulting data projection (right) of standardized flea data. The color and shape of data points are mapped to beetle species. The basis was identified using a projection pursuit guided tour, with the holes index. The contribution of the variables aede2 and tars1 approximately contrasts the other variables. The visible structure in the projection are the three clusters corresponding to the three species. Produced with the function \\code{view\\_basis()}."}
## Flea holes tour
set.seed(20190425)
f_dat  <- tourr::rescale(flea[,1:6])
f_cat  <- factor(flea$species)
## hard code a basis. tourr doesn't fix results with set.seed().
f_bas <-  c(.693, -.022, .082, -.119, .706, .023, 
             -.070, .438, .405, .515, .103, .604) %>% 
  matrix( ncol=2) %>% 
  tourr::orthonormalise()
f_mvar <- 5
f_msp  <- create_manip_space(basis = f_bas, manip_var = f_mvar)
f_proj <- data.frame(tourr::rescale(f_dat %*% f_msp[, 1:2]))

# step0, output
# Adjust centering to make axes and data side by side
oblique_frame(f_bas, data = f_dat, manip_var = f_mvar, lab = colnames(f_dat),
              color = f_cat, shape = f_cat) + 
  coord_fixed() + theme_void() + 
  theme(legend.position = "none")
```

@@

```{r step2, echo=F, out.width='70%',                                                fig.cap = "Illustration of a 3D manip space, this space is rotated effectively changing the contribution of the manip variable, aede2 in the example data. The blue circle and variable map lies on the projection plane. The red circle, orthogonal to the projection plane, illustrates the manipulation space and how the manip var can be controlled and how this affects the variable contribution back onto the projection plane. The other variables are omitted from the manipulation dimension for simplicity. Picturing the other variables in that dimension reveals the intuition that rotating one variable performs a constrained rotation on the others. This is illustrated with the \\code{view\\_manip\\_space()} function."}
view_manip_space2(basis = f_bas, manip_var = f_mvar, lab = colnames(f_dat))
```

@@

```{r step3, echo=F, warning=F, out.width='100%',  fig.cap = "Snapshots of a radial manual tour manipulating aede2: (1) original projection, (2) full contribution, (3) zero contribution, (4) back to original. "}
f_angle <- .29
f_mtour <- manual_tour(f_bas, manip_var = f_mvar, angle = f_angle)
if(dim(f_mtour)[3] != 15) message(
  paste0("!!! n_frames != 15 !!! Check angle step size. n_frames = ", dim(f_mtour)[3], ". angle = ", f_angle))

@@

```{r step4, echo=F, warning=F, out.height='7in', out.width='5.83in',                                           fig.cap = "Radial manual tour manipulating aede2 of standardized flea data. The axis for aede2 increases in contribution to the projection, from its initial value to 1, decreasing to 0 and then returning to the initial value. This effects the separation between the purple and green clusters. This shows that aede2 is important for distinguishing the purple species, because the separation disappears when aede2 is not contributing to the projection. An animation can be viewed at https://nspyrison.netlify.com/thesis/flea\\_manualtour\\_mvar5/."}
#play_manual_tour(f_dat, f_bas, f_mvar, col = f_cat, angle = f_angle)
array2ggfacets(tour_array = f_mtour, data = f_dat, m_var = f_mvar, class = f_cat)
```

@@

<!-- setup -->
```{r step0CODEEXAMPLE, eval=F, echo=T}
library(spinifex)
## Standardized flea data
f_data <- tourr::rescale(flea[, 1:6])
## Guided tour path, holes index
f_path <- save_history(f_data, guided_tour(holes()))
## Local extrema found
f_basis <- matrix(f_path[,, max(dim(f_path)[3])], ncol=2)
## Categorical class variable
f_cat <- factor(flea$species)
## Manip var, number of the variable to alter
f_mvar <- 5
## Anglular dist between frames (radians)
step_size <- .26
## Render and play animate, as plotly object by default
play_manual_tour(data = f_data,
                 basis = f_basis,
                 manip_var = f_mvar, 
                 angle = step_size,
                 col = f_cat,
                 pch = f_cat)
```

@@

```{r eval=FALSE}
## View a basis and projected data 
oblique_frame(basis = f_basis, 
              data = f_data,
              color = f_cat,
              shape = f_cat)
```

@@

jet_bas   <- JetCluster_basis
jet_dat   <- tourr::rescale(JetCluster_sub[, 1:4])
jet_cat   <- factor(JetCluster_sub$exp)
jet_mvar  <- 4
jet_ang   <- .315
jet_mtour <- manual_tour(basis = jet_bas, manip_var = jet_mvar, angle = jet_ang)
#dim(jet_mtour)

array2ggfacets(tour_array = jet_mtour, data = jet_dat,
             m_var = jet_mvar, class = jet_cat)
# play_manual_tour(data = jet_dat, basis = jet_bas, manip_var = jet_mvar,
#                  angle = jet_ang, axes = "bottomleft", col = jet_cat)

@@

```{r JetClusterBad, echo=F, warning=F, out.height='7in', out.width='5.83in',                                       fig.cap = "Snapshots of a radial manual tour of PC3 within the jet cluster, with color indicating experiment type: ATLAS7new (green) and ATLAS7old (orange).  When the contribution from PC3 is changed there is little change to the structure of the two groups, suggesting that PC3 is not important for distinguishing the experiments. The animation can be viewed at https://nspyrison.netlify.com/thesis/jetcluster\\_manualtour\\_pc3/."}
# Jet cluster, Fig 7 of cook_dynalical_2018, subset to ATLAS7old and ATLAS7new
jet_mvar  <- 3
jet_mtour <- manual_tour(basis = jet_bas, manip_var = jet_mvar, angle = jet_ang)

array2ggfacets(tour_array = jet_mtour, data = jet_dat,
             m_var = jet_mvar, class = jet_cat)
```

@@

```{r DISclusterGood, echo=F, warning=F, out.height='7in', out.width='5.83in',                                          fig.cap = "Snapshots of a radial manual tour exploring the sensitivity PC6 has on the structure of the DIS cluster, with color indicating experiment type: DIS HERA1+2 (green), dimuon SIDIS (purple), and charm SIDIS (orange). DIS HERA1+2 is distributed in a cross-shaped plane, charm SIDIS occupies the center of this cross, and dimuon SIDIS is a linear cluster crossing DIS HERA1+2. As the contribution of PC6 is increased, DIS HERA1+2 becomes almost singular in one direction (frame 5), indicating that this experiment has very little variability in the direction of PC6. The animation can be viewed at https://nspyrison.netlify.com/thesis/discluster\\_manualtour\\_pc6/."}

##DIS cluster, fig 8 of cook_dynamical_2018
load("./data/grDIScenter.rda")
load("./data/DIScluster_centered_basis.rda")
DIS_bas   <- DIScluster_centered_basis
DIS_dat   <- tourr::rescale(grDIScenter[, 1:6])
DIS_cat   <- factor(grDIScenter$disID)
DIS_mvar  <- 6
DIS_ang   <- .32
DIS_mtour <- manual_tour(basis = DIS_bas, manip_var = DIS_mvar, angle = DIS_ang)

array2ggfacets(tour_array = DIS_mtour, data = DIS_dat,
             m_var = DIS_mvar, class = DIS_cat)
# play_manual_tour(data = DIS_dat, basis = DIS_bas, manip_var = DIS_mvar,
#                  angle = DIS_ang, axes="bottomleft", col = DIS_cat)

@@

```{r DISclusterBad, echo=F, warning=F, out.height='7in', out.width='5.83in',                                       fig.cap = "Snapshots of a radial manual tour exploring the sensitivity PC2 to the structure of the DIS cluster, with color indicating experiment type: DIS HERA1+2 (green), dimuon SIDIS (purple), and charm SIDIS (orange). As contribution from PC2 is decreased, dimuon SIDIS becomes more distinguishable from the other two clusters (frames 10-14), indicating that in its absence PC2 is important. The animation can be viewed at https://nspyrison.netlify.com/thesis/discluster\\_manualtour\\_pc2/."}
##DIS cluster, fig 8 of cook_dynamical_2018
DIS_mvar  <- 2
DIS_mtour <- manual_tour(basis = DIS_bas, manip_var = DIS_mvar, angle = DIS_ang)

array2ggfacets(tour_array = DIS_mtour, data = DIS_dat,
             m_var = DIS_mvar, class = DIS_cat)
```

>>>>

---
title: > 
  spinifex: An R Package for Creating a Manual Tour of Low-dimensional Projections of Multivariate Data
author:
  - name: Nicholas Spyrison
    affiliation: Monash University
    address:
    - Faculty of Information Technology
    email:  nicholas.spyrison@monash.edu
  - name: Dianne Cook
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
abstract: >
  Dynamic low-dimensional linear projections of multivariate data collectively known as *tours* provide an important tool for exploring multivariate data and models. The R package \pkg{tourr} provides functions for several types of tours: grand, guided, little, local and frozen. Each of these can be viewed dynamically, or saved into a data object for animation. This paper describes a new package, \pkg{spinifex}, which provides a manual tour of multivariate data where the projection coefficient of a single variable is controlled. The variable is rotated fully into the projection, or completely out of the projection. The resulting sequence of projections can be displayed as an animation, with functions from either the \pkg{plotly} or \pkg{gganimate} packages. By varying the coefficient of a single variable, it is possible to explore the sensitivity of structure in the projection to that variable. This is particularly useful when used with a projection pursuit guided tour to simplify and understand the solution. The use of the manual tour is applied particle physics data to illustrate the sensitivity of structure in a projection to specific variable contributions.
keywords: exploratory data analysis, projection pursuit, high dimensional data, data visualization, statistical graphics, data science
preamble: >
  % preamble packages
link-citations: true
fig_caption: yes
toc: false
bibliography: spyrison-cook.bib
output: rticles::rjournal_article
  #bookdown::pdf_book ## alternative
---
```{r setup_paper, include=FALSE}
knitr::opts_chunk$set(
  fig.align  = "center",
  echo       = TRUE,
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  error      = FALSE,
  cache      = FALSE,
  cache.lazy = FALSE
)
if (F){ ## From reproducing, please use this dev ver:
  remotes::install_github("nspyrison/spinifex", 
                          ref = "dc8244c7e3f470fe3eee86f5c577c0dae21136c3")
}
library(tourr)
library(spinifex) ## Use dev version listed above!
library(ggplot2)
library(gridExtra)
library(dplyr)
```

@@

- $\textbf{X}$, the data, an $n \times p$ numeric matrix to be projected to 2D.
- $\textbf{B} = (B_1,~ B_2)$, any 2D orthonormal projection basis, $p \times 2$ matrix, describing the projection from $\mathbb{R}^p \Rightarrow \mathbb{R}^2$. This is called this the "original projection" because it is the starting point for the manual tour.
- $k$, is the index of the variable to manipulate, called the "manip var".

- $\textbf{e}$, a 1D basis vector of length $p$, with 1 in the $k$-th position and 0 elsewhere.
- $\textbf{M}$ is a $p \times 3$ matrix, defining the 3D subspace where data rotation occurs and is called the manip(ulation) space.
- $\textbf{R}$, the 3D rotation matrix, for performing unconstrained 3D rotations within the manip space, $\textbf{M}$.
- $\theta$, the angle of in-projection rotation, for example, on the reference axes; $c_\theta, s_\theta$ are its cosine and sine.
- $\phi$, the angle of out-of-projection rotation, into the manip space; $c_\phi, s_\phi$ are its cosine and sine. The initial value for animation purposes is $\phi_1$.
- $\textbf{U}$, the axis of rotation for out-of-projection rotation orthogonal to $\textbf{e}$.
- $\textbf{Y}$, the resulting projection of the data through the manip space, $\textbf{M}$, and rotation matrix, $\textbf{R}$.

@@

```{r step0, echo=F, results='hide', out.width='70%',                                 fig.cap="Initial 2D projection: representation of the basis (left) and resulting data projection (right) of standardized flea data. The color and shape of data points are mapped to beetle species. The basis was identified using a projection pursuit guided tour, with the holes index. The contribution of the variables aede2 and tars1 approximately contrasts the other variables. The visible structure in the projection are the three clusters corresponding to the three species. Produced with the function \\code{view\\_basis()}."}
## Flea holes tour
set.seed(20190425)
f_dat <- tourr::rescale(flea[,1:6])
f_clas <- factor(flea$species)
## hard code a basis. tourr doesn't fix results with set.seed().
f_bas <- c(.693, -.022, .082, -.119, .706, .023, 
           -.070, .438, .405, .515, .103, .604) %>% 
  matrix( ncol=2) %>% 
  tourr::orthonormalise()
f_mvar <- 5
f_msp <- create_manip_space(basis = f_bas, manip_var = f_mvar)
f_proj <- data.frame(tourr::rescale(f_dat %*% f_msp[, 1:2]))

# step0, output
# Adjust centering to make axes and data side by side
oblique_frame(f_bas, data = f_dat, manip_var = f_mvar, lab = colnames(f_dat),
              color = f_clas, shape = f_clas) + 
  coord_fixed() + theme_void() + 
  theme(legend.position = "none")
```

@@

```{r step2, echo=F, out.width='70%',                                                fig.cap="Illustration of a 3D manip space, this space is rotated effectively changing the contribution of the manip variable, aede2 in the example data. The blue circle and variable map lies on the projection plane. The red circle, orthogonal to the projection plane, illustrates the manipulation space and how the manip var can be controlled and how this affects the variable contribution back onto the projection plane. The other variables are omitted from the manipulation dimension for simplicity. Picturing the other variables in that dimension reveals the intuition that rotating one variable performs a constrained rotation on the others. This is illustrated with the \\code{view\\_manip\\_space()} function."}
view_manip_space2(basis = f_bas, manip_var = f_mvar, lab = colnames(f_dat))
```

@@

```{r step3, echo=F, warning=F, out.width='100%',                                     fig.cap="Snapshots of a radial manual tour manipulating aede2: (1) original projection, (2) full contribution, (3) zero contribution, (4) back to original. "}
f_angle <- .29
f_mtour <- manual_tour(f_bas, manip_var = f_mvar, angle = f_angle)
if(dim(f_mtour)[3] != 15) message(
  paste0("!!! n_frames != 15 !!! Check angle step size. n_frames = ", dim(f_mtour)[3], ". angle = ", f_angle))

@@

```{r step4, echo=F, warning=F, out.height='7in', out.width='5.83in',                 fig.cap="Radial manual tour manipulating aede2 of standardized flea data. The axis for aede2 increases in contribution to the projection, from its initial value to 1, decreasing to 0 and then returning to the initial value. This effects the separation between the purple and green clusters. This shows that aede2 is important for distinguishing the purple species, because the separation disappears when aede2 is not contributing to the projection. An animation can be viewed at https://nspyrison.netlify.com/thesis/flea\\_manualtour\\_mvar5/."}
## Figure
array2ggfacets(tour_array = f_mtour, data = f_dat, m_var = f_mvar, class = f_clas)

## Saving gif for the supplementary material:
if (FALSE){
  play_manual_tour(basis = f_bas, data = f_dat, manip_var = f_mvar,
                   color = f_clas, shape = f_clas,
                   ggtheme = list(theme_void(), 
                                  ggtitle("flea data, radial tour of aede2")),
                   fps = 8, start_pause = .5, end_pause = 1,
                   render_type = render_gganimate,
                   gif_filename = "flea_manualtour_mvar5.gif", 
                   gif_path = "./output")
}
```

@@

<!-- setup -->
```{r step0CODEEXAMPLE, eval=F, echo=T}
library(spinifex)
## Standardized flea data
f_data <- tourr::rescale(flea[, 1:6])
## Guided tour path, holes index
f_path <- save_history(f_data, guided_tour(holes()))
## Local extrema found
f_basis <- matrix(f_path[,, max(dim(f_path)[3])], ncol=2)
## Categorical class variable
f_clas <- factor(flea$species)
## Manip var, number of the variable to alter
f_mvar <- 5
## Anglular dist between frames (radians)
step_size <- .26
## Render and play animate, as plotly object by default
play_manual_tour(data = f_data,
                 basis = f_basis,
                 manip_var = f_mvar, 
                 angle = step_size,
                 col = f_clas,
                 pch = f_clas)
```

@@

```{r eval=FALSE}
## View a basis and projected data 
oblique_frame(basis = f_basis, 
              data = f_data,
              color = f_clas,
              shape = f_clas)
```

@@

jet_bas   <- JetCluster_basis
jet_dat   <- tourr::rescale(JetCluster_sub[, 1:4])
jet_clas   <- factor(JetCluster_sub$exp)
jet_mvar  <- 4
jet_ang   <- .315
jet_mtour <- manual_tour(basis = jet_bas, manip_var = jet_mvar, angle = jet_ang)

## Figure
array2ggfacets(tour_array = jet_mtour, data = jet_dat,
               m_var = jet_mvar, class = jet_clas)

## Saving gif for the supplementary material:
if (FALSE){
  play_manual_tour(basis = jet_bas, data = jet_dat, manip_var = jet_mvar,
                   color = jet_clas, shape = jet_clas,
                   ggtheme = list(theme_void(), 
                                  ggtitle("jet cluster, radial tour of PC4")),
                   fps = 8, start_pause = .5, end_pause = 1,
                   render_type = render_gganimate,
                   gif_filename = "jetcluster_manualtour_pc4.gif", 
                   gif_path = "./output")
}

@@

```{r JetClusterBad, echo=F, warning=F, out.height='7in', out.width='5.83in',         fig.cap="Snapshots of a radial manual tour of PC3 within the jet cluster, with color indicating experiment type: ATLAS7new (green) and ATLAS7old (orange).  When the contribution from PC3 is changed there is little change to the structure of the two groups, suggesting that PC3 is not important for distinguishing the experiments. The animation can be viewed at https://nspyrison.netlify.com/thesis/jetcluster\\_manualtour\\_pc3/."}
# Jet cluster, Fig 7 of cook_dynalical_2018, subset to ATLAS7old and ATLAS7new
jet_mvar  <- 3
jet_mtour <- manual_tour(basis = jet_bas, manip_var = jet_mvar, angle = jet_ang)

## Figure
array2ggfacets(tour_array = jet_mtour, data = jet_dat,
               m_var = jet_mvar, class = jet_clas)

## Saving gif for the supplementary material:
if (FALSE){
  play_manual_tour(basis = jet_bas, data = jet_dat, manip_var = jet_mvar,
                   color = jet_clas, shape = jet_clas,
                   ggtheme = list(theme_void(), 
                                 ggtitle("jet cluster, radial tour of PC3")),
                   fps = 8, start_pause = .5, end_pause = 1,
                   render_type = render_gganimate,
                   gif_filename = "jetcluster_manualtour_pc3.gif", 
                   gif_path = "./output")
}
```

@@

```{r DISclusterGood, echo=F, warning=F, out.height='7in', out.width='5.83in',        fig.cap="Snapshots of a radial manual tour exploring the sensitivity PC6 has on the structure of the DIS cluster, with color indicating experiment type: DIS HERA1+2 (green), dimuon SIDIS (purple), and charm SIDIS (orange). DIS HERA1+2 is distributed in a cross-shaped plane, charm SIDIS occupies the center of this cross, and dimuon SIDIS is a linear cluster crossing DIS HERA1+2. As the contribution of PC6 is increased, DIS HERA1+2 becomes almost singular in one direction (frame 5), indicating that this experiment has very little variability in the direction of PC6. The animation can be viewed at https://nspyrison.netlify.com/thesis/discluster\\_manualtour\\_pc6/."}

##DIS cluster, fig 8 of cook_dynamical_2018
load("./data/grDIScenter.rda")
load("./data/DIScluster_centered_basis.rda")
DIS_bas   <- DIScluster_centered_basis
DIS_dat   <- tourr::rescale(grDIScenter[, 1:6])
DIS_clas   <- factor(grDIScenter$disID)
DIS_mvar  <- 6
DIS_ang   <- .32
DIS_mtour <- manual_tour(basis = DIS_bas, manip_var = DIS_mvar, angle = DIS_ang)

## Figure
array2ggfacets(tour_array = DIS_mtour, data = DIS_dat,
               m_var = DIS_mvar, class = DIS_clas)

## Saving gif for the supplementary material
if (FALSE){
  play_manual_tour(basis = DIS_bas, data = DIS_dat, manip_var = DIS_mvar,
                   color = DIS_clas, shape = DIS_clas,
                   ggtheme = list(theme_void(), 
                                  ggtitle("DIS cluster, radial tour of PC6")),
                   fps = 8, start_pause = .5, end_pause = 1,
                   render_type = render_gganimate,
                   gif_filename = "discluster_manualtour_pc6.gif", 
                   gif_path = "./output")
}

@@

```{r DISclusterBad, echo=F, warning=F, out.height='7in', out.width='5.83in',         fig.cap="Snapshots of a radial manual tour exploring the sensitivity PC2 to the structure of the DIS cluster, with color indicating experiment type: DIS HERA1+2 (green), dimuon SIDIS (purple), and charm SIDIS (orange). As contribution from PC2 is decreased, dimuon SIDIS becomes more distinguishable from the other two clusters (frames 10-14), indicating that in its absence PC2 is important. The animation can be viewed at https://nspyrison.netlify.com/thesis/discluster\\_manualtour\\_pc2/."}
##DIS cluster, fig 8 of cook_dynamical_2018
DIS_mvar  <- 2
DIS_mtour <- manual_tour(basis = DIS_bas, manip_var = DIS_mvar, angle = DIS_ang)

## Figure
array2ggfacets(tour_array = DIS_mtour, data = DIS_dat,
               m_var = DIS_mvar, class = DIS_clas)

## Saving gif for the supplementary material
if (FALSE){
  play_manual_tour(basis = DIS_bas, data = DIS_dat, manip_var = DIS_mvar,
                   color = DIS_clas, shape = DIS_clas,
                   ggtheme = list(theme_void(), 
                                  ggtitle("jet cluster, radial tour of PC2")),
                   fps = 8, start_pause = .5, end_pause = 1,
                   render_type = render_gganimate,
                   gif_filename = "discluster_manualtour_pc2.gif", 
                   gif_path = "./output")
}
```

### 2020-08-25 paper/spyrison-cook.tex
caching issue, view_basis to oblique_frame
Code touch up, gifs saved to /output/.

\begin{itemize}
\tightlist
\item
  \(\textbf{X}\), the data, an \(n \times p\) numeric matrix to be
  projected to 2D.
\item
  \(\textbf{B} = (B_1,~ B_2)\), any 2D orthonormal projection basis,
  \(p \times 2\) matrix, describing the projection from
  \(\mathbb{R}^p \Rightarrow \mathbb{R}^2\). This is called this the
  ``original projection'' because it is the starting point for the
  manual tour.
\item
  \(k\), is the index of the variable to manipulate, called the ``manip
  var''. 
\item
  \(\textbf{e}\), a 1D basis vector of length \(p\), with 1 in the
  \(k\)-th position and 0 elsewhere.
\item
  \(\textbf{M}\) is a \(p \times 3\) matrix, defining the 3D subspace
  where data rotation occurs and is called the manip(ulation) space.
\item
  \(\textbf{R}\), the 3D rotation matrix, for performing unconstrained
  3D rotations within the manip space, \(\textbf{M}\).
\item
  \(\theta\), the angle of in-projection rotation, for example, on the
  reference axes; \(c_\theta, s_\theta\) are its cosine and sine.
\item
  \(\phi\), the angle of out-of-projection rotation, into the manip
  space; \(c_\phi, s_\phi\) are its cosine and sine. The initial value
  for animation purposes is \(\phi_1\).
\item
  \(\textbf{U}\), the axis of rotation for out-of-projection rotation
  orthogonal to \(\textbf{e}\).
\item
  \(\textbf{Y}\), the resulting projection of the data through the manip
  space, \(\textbf{M}\), and rotation matrix, \(\textbf{R}\).
\end{itemize}

@@

\begin{Schunk}
\begin{Sinput}
library(spinifex)
## Standardized flea data
f_data <- tourr::rescale(flea[, 1:6])
## Guided tour path, holes index
f_path <- save_history(f_data, guided_tour(holes()))
## Local extrema found
f_basis <- matrix(f_path[,, max(dim(f_path)[3])], ncol=2)
## Categorical class variable
f_cat <- factor(flea$species)
## Manip var, number of the variable to alter
f_mvar <- 5
## Anglular dist between frames (radians)
step_size <- .26
## Render and play animate, as plotly object by default
play_manual_tour(data = f_data,
                 basis = f_basis,
                 manip_var = f_mvar, 
                 angle = step_size,
                 col = f_cat,
                 pch = f_cat)
\end{Sinput}
\end{Schunk}

@@

The function \texttt{view\_basis} can be used to show a projection of
the basis, or with the data overlaid. For example, the plots in Figures
\ref{fig:step0} and \ref{fig:step3} were made with code similar to this:

\begin{Schunk}
\begin{Sinput}
## View a basis and projected data 
oblique_frame(basis = f_basis, 
              data = f_data,
              color = f_cat,
              shape = f_cat)
\end{Sinput}
\end{Schunk}

>>>>

\begin{itemize}
\item
  \(\textbf{X}\), the data, an \(n \times p\) numeric matrix to be
  projected to 2D.
\item
  \(\textbf{B} = (B_1,~ B_2)\), any 2D orthonormal projection basis,
  \(p \times 2\) matrix, describing the projection from
  \(\mathbb{R}^p \Rightarrow \mathbb{R}^2\). This is called this the
  ``original projection'' because it is the starting point for the
  manual tour.
\item
  \(k\), is the index of the variable to manipulate, called the ``manip
  var''.
\item
  \(\textbf{e}\), a 1D basis vector of length \(p\), with 1 in the
  \(k\)-th position and 0 elsewhere.
\item
  \(\textbf{M}\) is a \(p \times 3\) matrix, defining the 3D subspace
  where data rotation occurs and is called the manip(ulation) space.
\item
  \(\textbf{R}\), the 3D rotation matrix, for performing unconstrained
  3D rotations within the manip space, \(\textbf{M}\).
\item
  \(\theta\), the angle of in-projection rotation, for example, on the
  reference axes; \(c_\theta, s_\theta\) are its cosine and sine.
\item
  \(\phi\), the angle of out-of-projection rotation, into the manip
  space; \(c_\phi, s_\phi\) are its cosine and sine. The initial value
  for animation purposes is \(\phi_1\).
\item
  \(\textbf{U}\), the axis of rotation for out-of-projection rotation
  orthogonal to \(\textbf{e}\).
\item
  \(\textbf{Y}\), the resulting projection of the data through the manip
  space, \(\textbf{M}\), and rotation matrix, \(\textbf{R}\).
\end{itemize}

@@

\begin{Schunk}
\begin{Sinput}
library(spinifex)
## Standardized flea data
f_data <- tourr::rescale(flea[, 1:6])
## Guided tour path, holes index
f_path <- save_history(f_data, guided_tour(holes()))
## Local extrema found
f_basis <- matrix(f_path[,, max(dim(f_path)[3])], ncol=2)
## Categorical class variable
f_clas <- factor(flea$species)
## Manip var, number of the variable to alter
f_mvar <- 5
## Anglular dist between frames (radians)
step_size <- .26
## Render and play animate, as plotly object by default
play_manual_tour(data = f_data,
                 basis = f_basis,
                 manip_var = f_mvar, 
                 angle = step_size,
                 col = f_clas,
                 pch = f_clas)
\end{Sinput}
\end{Schunk}

@@

The function \texttt{oblique\_frame} can be used to show a projection of
the basis, or with the data overlaid. For example, the plots in Figures
\ref{fig:step0} and \ref{fig:step3} were made with code similar to this:

\begin{Schunk}
\begin{Sinput}
## View a basis and projected data 
oblique_frame(basis = f_basis, 
              data = f_data,
              color = f_clas,
              shape = f_clas)
\end{Sinput}
\end{Schunk}
